---
title: 'Section 1: Setup, Quality Control and Sample Integration'
teaching: 10
exercises: 2
---


:::::::::::::::::::::::::::::::::::::: questions 

- hello!

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- hello!

::::::::::::::::::::::::::::::::::::::::::::::::

### Step 1. Load the packages and data

Today we'll be working with Seurat (a popular scRNA-seq analysis
package). SeuratData will be used to load in the experimental data we're
analysing. Tidyverse is a fundamental and very popularly used set of
tools to wrangle and visualise data.

We'll need to load the DESeq2 R package for when we explore pseudobulk
DE approaches

pheatmap and grid are two really useful packages for creating custom
heatmaps with our scRNA-seq data and exporting figures, respectively.

``` {r, eval=FALSE,message=FALSE,warning=FALSE}
library(Seurat)
library(SeuratData)
library(tidyverse)
library(DESeq2)
library(patchwork)
library(pheatmap)
library(grid)
library(metap)

set.seed(4242) # Set Seed for Reproducibility
```


``` {r, echo=FALSE,message=FALSE,warning=FALSE}
library(Seurat)
#library(SeuratData)
library(tidyverse)
library(DESeq2)
library(patchwork)
library(pheatmap)
library(grid)
library(metap)

set.seed(4242) # Set Seed for Reproducibility
```

We're using the ifnb public dataset provided by Seurat. This dataset
contains PBMC data from 8 lupus patients before and after interferon
beta treatment.

I strongly encourage you to explore the other datasets offered by the
SeuratData package, it can be really good practice in your spare time.

The ifnb Seurat object we're loading in here was originally made in
Seurat v4, there have since been a lot of changes from Seurat v4 to v5
so we'll use the `UpdateSeuratObject()` function to update the Seurat
object so that it is compatible for today.


``` {r, eval=FALSE}
head(AvailableData()) # if you want to see the available SeuratData datasets use this function
```


``` {r, eval=FALSE}
InstallData("ifnb") # install our treatment vs control dataset for today

data("ifnb") # Load the dataset into our current R script
ifnb <- UpdateSeuratObject(ifnb) # Make sure the seurat object is in the format of Seurat v5

str(ifnb) # we can use this to take a look at the information in our Seurat Object
```


```{r,echo=FALSE}

load('data/ifnb.RData')

```


::::::::::::::::::::::::::::::::::::: challenge 

Looking at the output from the `str()` function above, can
you tell whether this seurat object is processed or unprocessed?

:::::::::::::::::::::::: solution 

When loading in seurat objects, we can have a look at what processing steps have been performed on it by using the str() function. In the output we can tell that the ifnb Seurat object is unprocessed because the scale.data slot is empty, no variable features have been identified, and no dimensionality reduction functions have been performed.  

:::::::::::::::::::::::::::::::::


:::::::::::::::::::::::::::::::::::

### Step 2: Run QC, filter out low quality cells

Lets start by processing our data (run the standard seurat workflow
steps including preprocessing and filtering).

First we need to take a look at QC metrics, then decide on the
thresholds for filtering.


:::::::: discussion

Looking at the violin plots of QC metrics, what do you
think about the overall quality of the ifnb dataset?

::::::::


``` {r}
# Step 2a: QC and filtering
ifnb$percent.mt <- PercentageFeatureSet(object = ifnb, pattern = "^MT-") # First let's annotate the mitochondrial percentage for each cell

head((ifnb@meta.data)) # we can take a look mitochondrial percentages for the seurat object by viewing the seurat objects metadata

# Step 2b: Visualise QC metrics and identify filtering thresholds
qc.metric.plts <- VlnPlot(ifnb, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3) + 
  ggtitle("Before Filtering")

association.plt.raw <- FeatureScatter(ifnb, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + geom_smooth(method = "lm") +
  ggtitle("Before Filtering")

qc.metric.plts
association.plt.raw
```


After visualising QC metrics, we'll move on to the actual filtering

``` {r}
# Step 2c: filter out low-quality cells + visualise the metrics for our filtered seurat object
ifnb.filtered <- subset(ifnb, subset = nCount_RNA > 800 & 
                          nCount_RNA < 5000 &
                          nFeature_RNA > 200 &
                          nFeature_RNA < 1200 &
                          percent.mt < 5)

qc.metric.plts.filtered <- VlnPlot(ifnb.filtered, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3) +
  ggtitle("After Filtering")
association.plt.filtered <- FeatureScatter(ifnb.filtered, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + geom_smooth(method = "lm") +
  ggtitle("After Filtering")

qc.metric.plts.filtered
association.plt.filtered
```

Let's check how many cells we've filtered out (looks like \~400 cells
were removed):

``` {r}
## Defining a couple helper functions to standardise x and y axis for two plots
get_plot_range <- function(plot) {
  data <- layer_data(plot)
  list(
    x = range(data$x, na.rm = TRUE),
    y = range(data$y, na.rm = TRUE)
  )
}
standardise_plt_scale <- function(plt1, plt2){

  # Get ranges for both plots
  range_raw <- get_plot_range(plt1)
  range_filtered <- get_plot_range(plt2)
  
  # Calculate overall range
  x_range <- range(c(range_raw$x, range_filtered$x))
  y_range <- range(c(range_raw$y, range_filtered$y))
  
  suppressMessages({
  # Update both plots with the same x and y scales
  association.plt.raw <- association.plt.raw +
    scale_x_continuous(limits = x_range) +
    scale_y_continuous(limits = y_range)
  
  association.plt.filtered <- association.plt.filtered +
    scale_x_continuous(limits = x_range) +
    scale_y_continuous(limits = y_range)
  })
  
  # Wrap the plots
  wrapped_plots <- wrap_plots(list(association.plt.raw, association.plt.filtered), 
                              ncol = 2)

  return(wrapped_plots)
}

wrap_plots(list(qc.metric.plts, qc.metric.plts.filtered), 
           ncol = 1)
```


``` {r}
association.plts <- standardise_plt_scale(association.plt.raw,
                                          association.plt.filtered)
association.plts
```


Let's check how many cells we've filtered out (looks like \~400 cells
were removed):

``` {r}
ifnb
```


``` {r}
ifnb.filtered
```



Next we need to split our count matrices based on conditions. This step
stores stimulated versus unstimulated expression information separately,
creating a list of RNA assays grouped by the "stim" condition. Note:
this is important for downstream integration steps in Seurat v5.

``` {r}
ifnb.filtered[["RNA"]] <- split(ifnb.filtered[["RNA"]], f = ifnb.filtered$stim) # Lets split our count matrices based on conditions (stored within different layers) -> needed for integration steps in Seurat v5
```

### Step 3: Before performing differential expression between the two conditions, let's assess whether we need to integrate our data

After filtering out low quality cells, we want to visualise our data to
see how cells group by condition and if we need to perform batch-effect
correction (integration)

``` {r}
ifnb.filtered <- NormalizeData(ifnb.filtered)

ifnb.filtered <- FindVariableFeatures(ifnb.filtered)

ifnb.filtered <- ScaleData(ifnb.filtered)

## Centering and scaling data matrix

ifnb.filtered <- RunPCA(ifnb.filtered)



ElbowPlot(ifnb.filtered) # Visualise the dimensionality of the data, looks like 15 PCs is adequate to capture the majority of the variation in the data, but we'll air on the higher side and consider all 20 dimensions.
```



``` {r}
ifnb.filtered <- RunUMAP(ifnb.filtered, dims = 1:20, reduction = 'pca')


DimPlot(ifnb.filtered, reduction = 'umap', group.by = 'stim') # lets see how our cells separate by condition and whether integration is necessary
```



``` {r}
DimPlot(ifnb.filtered, reduction = 'pca', group.by = 'stim') # lets see how our cells separate by condition and whether integration is necessary
```



These are PBMCs before and after treatment, there should be cells that
are similar between both conditions, it looks like we'll have to run
some batch effect correction to overlay similar cell-types from both
conditions to perform downstream analysis.


:::: discussion

Do you think we need to integrate our data? 
Hint: Look at the UMAP and PC1/PC2 plots we made above 

What do you think would
happen if we were to perform unsupervised clustering right now, without
integrating our data (or overlaying similar cells on top of each other
from both conditions)?


::::


### Step 4: Integrating our data using the harmony method

Seurat v5 has made it really easy to test different integration methods
quickly, let's use a really popular approach (harmony) first.

``` {r}
# code adapted from: https://satijalab.org/seurat/articles/seurat5_integration
ifnb.filtered <- IntegrateLayers(object = ifnb.filtered,
                                 method = HarmonyIntegration,
                                 orig.reduction = "pca", 
                                 new.reduction = "harmony")


ifnb.filtered <- RunUMAP(ifnb.filtered, reduction = "harmony", dims = 1:20, reduction.name = "umap.harmony")

after.harmony <- DimPlot(ifnb.filtered, reduction = "umap.harmony", group.by = "stim") + 
  ggtitle("After Harmony Integration")
before.integration <- DimPlot(ifnb.filtered, reduction = "umap", group.by = "stim") +
  ggtitle("Before Integration")

before.integration | after.harmony
```


:::: discussion

Looking at the UMAPs above, do you think integration was
successful? Have a slide on what if its just different cell types. !!!
question Try looking at the PC1 and PC2 plots for harmony and seurat as
well

::::



### Step 5: Integrating our data using an alternative Seurat CCA
method

``` {r}
ifnb.filtered <- IntegrateLayers(object = ifnb.filtered,
                                 method = CCAIntegration,
                                 orig.reduction = "pca", 
                                 new.reduction = "integrated.cca")

ifnb.filtered <- RunUMAP(ifnb.filtered, reduction = "integrated.cca", dims = 1:20, reduction.name = "umap.cca")

after.seuratCCA <- DimPlot(ifnb.filtered, reduction = "umap.cca", group.by = "stim") +
  ggtitle("After Seurat CCA Integration")

before.integration | after.seuratCCA
```



``` {r}
after.harmony | after.seuratCCA

## Show example slide of integration 'failing' but due to different cell types in each sample ***
```



:::::: discussion

What do you think of the integration results now?

:::::::

        
**Hint:** Also look at the PC1 and PC2 plots for each integration method.


### Step 6: Perform standard clustering steps after integration

This step collapses individual control and treatment datasets together
and needs to be done before differential expression analysis

``` {r}
ifnb.filtered <- FindNeighbors(ifnb.filtered, reduction = "integrated.cca", dims = 1:20)



ifnb.filtered <- FindClusters(ifnb.filtered, resolution = 0.5)



ifnb.filtered <- JoinLayers(ifnb.filtered)
```


::::::::::::::::::::::::::::::::::::: keypoints 

- 

::::::::::::::::::::::::::::::::::::::::::::::::

